---
title: "Reproducible Research: Peer Assessment 1"
author: "John Snell"
date: "17 October 2015"
output: 
  html_document:
    keep_md: true
---


#Introduction
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self" movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.  


#Data

The data for this assignment can be downloaded from the course web site:

Dataset: Activity monitoring data [52K]
The variables included in this dataset are:   
**steps**: Number of steps taking in a 5-minute interval (missing values are coded as NA)  
**date**: The date on which the measurement was taken in YYYY-MM-DD format  
**interval**: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of **17,568** observations in this dataset.   

#Assignment

This assignment will be described in multiple parts. You will need to write a report that answers the questions detailed below. Ultimately, you will need to complete the entire assignment in a single R markdown document that can be processed by knitr and be transformed into an HTML file.

Throughout your report make sure you always include the code that you used to generate the output you present. When writing code chunks in the R markdown document, always use echo = TRUE so that someone else will be able to read the code. This assignment will be evaluated via peer assessment so it is essential that your peer evaluators be able to review the code for your analysis.

For the plotting aspects of this assignment, feel free to use any plotting system in R (i.e., base, lattice, ggplot2)

*Fork/clone the GitHub repository created for this assignment.*
You will submit this assignment by pushing your completed files into your forked repository on GitHub. The assignment submission will consist of the URL to your GitHub repository and the SHA-1 commit ID for your repository state.

NOTE: The GitHub repository also contains the dataset for the assignment so you do not have to download the data separately.

Loading and preprocessing the data

Show any code that is needed to

## Loading and preprocessing the data
*Load the data (i.e. read.csv())*

```{r}
##setwd(".\\RepData_PeerAssessment1")
activity<-read.csv("activity.csv")
```

Process/transform the data (if necessary) into a format suitable for your analysis.   


##What is mean total number of steps taken per day?

load the dplyr library to group the raw data on a per day basis.

```{r}
library(dplyr)
activity_per_day<-group_by(activity,date)
```

For this part of the assignment, missing values in the dataset are ignored.

###Calculate the total number of steps taken per day.

Using the summarise function from dplyr on the grouped data.

```{r}
sumsteps<-summarise(activity_per_day, sumstep = sum(steps, na.rm=TRUE))
print.data.frame(sumsteps)
```

###Presenting the total number of steps taken each day as a histogram.


load the ggplot functionality and use qplot to plot the histogram.

```{r}
library(ggplot2)
qplot(sumstep,data=sumsteps,xlab = "number of steps per day", ylab="frequency", 
      main="Histogram of Number of Steps Per Day", binwidth=1000)
```

###Calculate and report the mean and median of the total number of steps taken per day

```{r}
summarise(sumsteps,"mean steps per day"=mean(sumstep),
          "median steps per day"=median(sumstep))
```

##What is the average daily activity pattern?

###Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) 
###and the average number of steps taken, averaged across all days (y-axis)
*modify activity dataframe by adding a column by converting interval to five minute timed interval.*
```{r}
activityintervals<-activity$interval
activity$fivemininterval<-substr(as.POSIXct(sprintf("%04.0f", activityintervals), 
                                            format='%H%M'), 12, 16)
```

*group by 5 minutes timed interval.*
```{r}
stepsbyinterval<-group_by(activity,interval)
```

*establish average of each 5 minute slot across all days*
```{r}
averagestepsbyinterval<-summarise(stepsbyinterval,
                                  averagestepsforinterval=mean(steps,na.rm = TRUE))
print.data.frame(averagestepsbyinterval)
qplot(interval, averagestepsforinterval,data = averagestepsbyinterval, geom='line',xlab="interval",ylab="Average steps for 5 minute interval",
      main="Plot of average number of steps for each five minute interval")
```

###Which 5-minute interval, on average across all the days in the dataset, 
###contains the maximum number of steps?

```{r}
 averagestepsbyinterval[which.max(averagestepsbyinterval$averagestepsforinterval),]
```

##Imputing missing values.

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

###Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs).
```{r}
sum(is.na(activity))
```

Devise a strategy for filling in all of the missing values in the dataset. The 
strategy does not need to be sophisticated. For example, you could use the 
mean/median for that day, or the mean for that 5-minute interval, etc.

make a copy of the activity matrix
group by fiveminuteinterval
fill in na's based on the average steps for that specific time interval
```{r}
filledactivity<-activity
filledactivity<-group_by(filledactivity,fivemininterval)
averagedintervalsforfilledactivity<-summarise(filledactivity,
                                        meanintervalstep=mean(steps,na.rm=TRUE))
```
*Create a new dataset that is equal to the original dataset but with the missing* 
*data filled in.*

```{r}
#this algorithm steps through the activity data frame and wherever there is an NA 
#in the steps column, it fills it in with the average value of that interval, where
#the average is taken across all days for which there is data.
#
i = 1
while (i <= nrow(filledactivity))
{
        j=0
        while (j < nrow(averagedintervalsforfilledactivity)) 
                #number of rows corresponds to the numer of intervals in a day
        {
                if (is.na(filledactivity$steps[i + j]))
                {
                        filledactivity$steps[i + j] <-
                        averagedintervalsforfilledactivity$meanintervalstep[j+1]
                }
                j = j+ 1 #adjust the inner loop counter
        }
        
        i = i + j #adjust the out loop counter
}
```

*Make a histogram of the total number of steps taken each day.*

```{r}
filledactivity<-group_by(filledactivity,date)
totalstepsperday<-summarise(filledactivity,totalsteps = sum(steps))
print.data.frame(totalstepsperday)
qplot(totalsteps,data = totalstepsperday, binwidth = 1000, xlab="total steps per day",
      ylab = "frequency", main = "Histogram of total steps per day")
```

*Calculate and report the mean and median total number of steps taken per day.*
```{r}
summarise(totalstepsperday,"mean steps per day"=mean(totalsteps),"median steps per day"=median(totalsteps))
```

*Do these values differ from the estimates from the first part of the assignment?*

**Yes - both mean and median have increased in value.**

*What is the impact of imputing missing data on the estimates of the total daily number of steps?*

**Both the mean and median values are increased relative to the first part of the** 
**assignment where blank data was simply omitted.**
**The mean and the median are now the same value which fits in with the shape of **
**the histogram which is consderably more symmetrical than the histogram in the **
**first part of the assignment. Both the mean and median have increased in value **
**relative to the first part as would be expected since we are injecting values **
**into the dataset.**

##Are there differences in activity patterns between weekdays and weekends?

For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

*Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.*
```{r}
#create a new variable in filledactivity and fill it by detecting whether the 
#row date for this interval is a weekend or a weekday. 
filledactivity$daytype<-ifelse(weekdays(as.Date(filledactivity$date)) %in% c("Saturday","Sunday"),"weekend","weekday")

#convert the new column from character to factor
filledactivity$daytype<-as.factor(filledactivity$daytype)
```


*Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.*   
```{r}
#two layers of grouping - intially by type of day and then by interval
groupedfilledactivity<-group_by(filledactivity, daytype,interval)
#produce summary data which is the average number of steps for an interval across
#all days of a specfic day type
summarydata<-summarise(groupedfilledactivity, averagestepsbyinterval=mean(steps))
#present results
print.data.frame(summarydata)

#create a ggplot object

p<-ggplot(summarydata, aes(interval, averagestepsbyinterval)) + geom_line() + 
                        xlab("interval") + ylab("Average steps for interval")
#plot using panels based on type of day - panels are presented column wise
p+facet_grid(daytype~.)
```

*Are there differences in activity patterns between weekdays and weekends?*

**From the summary date it can be seen that activity starts later at weekends and**
**goes on later. The peak of the activity around 08:35 is fundamentally at the **
**same time but is less intense indicating a more relaxed appoach to the activity**
**which is now spread more over the subsequent part of the day.**

